\mytitle

### 2011-12-22 papers/PLDI2012/response_PLDI2012.txt

"the title is fairly vague and non optimal -- perhaps
"succinct" should be replaced by "disjunctive"?"
Path Focusing technique and our combined technique (with or without disjunctive
invariants) do not work on the classical control flow graph, but on a multigraph
where all the paths inside loops are distinguished. The term "succint" refers to
the fact we actually do not compute all the paths of this multigraph, but work
on an succint representation thereof (using an SMT formula), thus avoiding the
enumeration of an exponential number of paths in a huge number of cases.

"The results are unsurprising"
The experimental results show that techniques that distinguish paths inside a
program give more precise results, but, more importantly, at a reasonable cost
(which was not evident).
It is also interesting to see that more complex technique perform sometimes
faster that more naive approaches.

"The problem with path focusing is that (1) it will process all
executable paths at once, including potentially many redundant paths
that don't contribute to the solution; and (2) its widening operation
lacks the precision conferred by guided static analysis."

This is not exactly the point. The problem with path focusing is that it will compute
the fixpoint iterations on the whole graph at once, using classical widening
operations in order to guarantee the convergence.
When the fixpoint has been reached, it finally apply some narrowing iterations
to recover precision.
The idea here is to apply some narrowing iterations during the analysis, in
order to recover some precision before continuing.
To clarify, widening is the same, but we perform regularly narrowing iterations,
and not only at the end of the ascending iterations.

"how much benefit this technique confers to other types of abstract
domains, including those that don't require widening"
Intuitively, Guided Static Analysis technique performs better because it
recovers precision lost by widening. It will probably not increase the precision
when using domains without widening/narrowing. However, the iteration strategy
will be changed, so the computation time may differ compared to a classical
abstract interpretation. We can do the same remarks on our combined technique
compared to Path Focusing.

"I'd like to know more about how applicable the technique
is to other kinds of abstract domains"
The general idea does not depend on the abstract domain. However, we have to
transform an abstract value into an SMT formula. Such transformation is easy in
our case, since it is a cunjunction of linear inequalities, but may still
require some research for other abstract domains that could require other
theories (for instance, arrays could be
encoded in SMT using uninterpreted functions, etc.)

"2 to 5% difference, does it make a difference at all?"
Our prototype of analyzer performs many over-approximations (no memory model,
etc.), thus leading to invariants that are often very unprecise.
Intuitively, I would say we could have obtained more difference between the
different techniques by reducing the number of over-approximation we do for the
time being.

### 2012-02-09 papers/SAS2012/implicitization.tex
tassage

\documentclass[preprint]{sigplanconf}

% The following \documentclass options may be useful:
%
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% authoryear    To obtain author/year citation style instead of numeric.

%\newcommand{\MM}[1]{{\color{blue} TODO(MM): #1}}
%\newcommand{\JH}[1]{{\color{teal} TODO(JH): #1}}
%\newcommand{\DM}[1]{{\color{violet} TODO(DM): #1}}
%\newcommand{\MM}[1]{}
%\newcommand{\JH}[1]{}
%\newcommand{\DM}[1]{}

@@

\usepackage{amsmath}
\usepackage{mathcomp,amsfonts,amssymb,MnSymbol}
\usepackage{dmnatbib}
\usepackage[mathscr]{euscript}
\usepackage{algorithmic,algorithm}
%\usepackage{graphicx}
%\usepackage{color}

@@

\usepackage[pdfusetitle]{hyperref}

@@

@@

\begin{document}

\conferenceinfo{PLDI '12}{11-16 June 2012, Beijing} 
\copyrightyear{2012} 
\copyrightdata{[to be supplied]}

%\titlebanner{banner above paper title}        % These are ignored unless
\preprintfooter{Succinct representations for abstract interpretation}   % 'preprint' option specified.

\title{Succinct Representations for Abstract Interpretation}
\subtitle{Combined analysis algorithms and experimental evaluation}

\authorinfo{Anonymous}
           {The Internet}
           {anonymous.gal@gmail.com}

\maketitle

@@

% \category{CR-number}{subcategory}{third-level}
\category{D.2.4}{Software/Program Verification}{Formal methods}
\category{F.3.1}{Specifying and Verifying and Reasoning about Programs}{Assertions, Invariants, Mechanical verification, Pre- and post-conditions}
\category{F.3.2}{Semantics of Programming Languages}{Program analysis}
\category{F.4.1}{Mathematical Logic}{Logic and constraint programming, Mechanical theorem proving}
\category{I.2.3}{Deduction and Theorem Proving}{Inference engines, Logic programming, Resolution}

\terms
theory, verification, algorithms

\keywords
SMT-solving, static analysis, abstract interpretation, disjunctions

@@

This article makes the following contributions:
\begin{enumerate}
\item We recast the guided static analysis technique from \citet{DBLP:conf/sas/GopanR07} on the expanded multigraph from \citet{Monniaux_Gonnord_SAS11}, considering entire paths instead of individual transitions, using SMT queries and binary decision diagrams.\label{contr:guided_multigraph} (See \S\ref{sec:guided_multigraph})
\item We improve the technique for obtaining disjunctive invariants from \citet{DBLP:conf/pldi/GulwaniZ10} by replacing the explicit exhaustive enumeration of paths by a sequence of SMT queries.\label{contr:disjunctive} (See \S\ref{sec:disjunctive})
\item We implemented these techniques, in addition to ``classical'' iterations and the original guided static analysis, inside a prototype static analyzer.
This tool uses the LLVM bitcode format \cite{Lattner:2004:LCF:977395.977673,LLVM_langref} as input, which can be produced by compilation from C, C++ and Fortran, enabling it to be run on many real-life programs.
With respect to abstract domains on numerical variables, it uses the APRON library \cite{DBLP:conf/cav/JeannetM09}, which supports a variety of abstract domains from which we can choose with minimal changes to our analyzer.

@@

The expected inductive invariant is $\lstinline|x_old| \in [-100000,100000]$, but classical abstract interpretation using intervals (or octagons or polyhedra) finds $\lstinline|x_old| \in (-\infty,+\infty)$~\citep{ASTREE_ESOP05}.
Let us briefly see why.

@@

We could use ``widening up to'' or ``widening with thresholds'', propagating the ``magic values'' $\pm 100000$ associated to \lstinline|x| into \mbox{\lstinline|x_old|,} but these syntactic approaches cannot directly cope with programs for which $\lstinline|x|  \in [-100000,+100000]$ is itself obtained by analysis.
The guided static analysis of \citet{DBLP:conf/sas/GopanR07} does not perform
better, and also obtains $\lstinline|x_old| \in (-\infty,+\infty)$.

@@

Following recent suggestions \cite{Gawlitza_Monniaux_ESOP11,Monniaux_Gonnord_SAS11}, we avoid this space explosion by keeping the second program implicit while simulating its analysis. This means we work on an implicitly represented transition multigraph (Fig.~\ref{fig:multigraph}); it is succinctly representated by the transition graph of the first program.
Our first contribution (\S\ref{sec:guided_multigraph}) is to recast the ``guided analysis'' from \citet{DBLP:conf/sas/GopanR07} on such a succinct representation of the paths in lieu of the individual transitions.
A similar explosion occurs in disjunctive invariant generation, following \citet{DBLP:conf/pldi/GulwaniZ10}; our second contribution (\S\ref{sec:disjunctive}) applies our implicit representation to their method.

\subsection{Guided static analysis}
\label{sec:guided}
\emph{Guided static analysis} was proposed by \citet{DBLP:conf/sas/GopanR07} as an improvement over classical upward Kleene iterations with widening.
Consider the following program, taken from \cite{DBLP:conf/sas/GopanR07}:
\lstinputlisting[label=lst:gopan_reps]{gopan_reps.c}

\begin{figure}
\begin{center}
\includegraphics[scale=0.75]{figures/gopan_reps}
\end{center}
\caption{The invariant for Prog.~\ref{lst:gopan_reps}: the piecewise linear, solid line is the strongest invariant, the grayed polyhedron is its convex hull.}
\label{fig:gopan_reps_invariant} 
\end{figure}

@@

\citeauthor{DBLP:conf/sas/GopanR07}' idea is to analyze the first phase of the loop with a widening and narrowing sequence, and thus obtain $0 \leq x \leq 50 \land y = x$, and then analyze the second phase, finally obtaining invariant~\ref{eqn:triangle}; each phase is identified by the tests taken or not taken.

@@

\citet{Monniaux_Gonnord_SAS11}'s \emph{path-focusing} technique
distinguishes the different paths in the program in order to avoid loss of
precision due to merge operations. Since the number of paths may be exponential,
the technique keeps them implicit and computes them when needed using
SMT-solving.
The (accelerated) Kleene iterations (\S\ref{sec:static_analysis}) are computed over a reduced multigraph instead of the classical transition graph.

@@

Guided static analysis, as proposed by \citet{DBLP:conf/sas/GopanR07}, applies to the transition
graph of the program. We now present a new technique applying this analysis on the implicit
multigraph from \cite{Monniaux_Gonnord_SAS11}, thus avoiding control flow merges with
unfeasible paths.
In this section, we use the same notations as \ref{sec:path_focusing}, except
that we call the abstract values $X_i^s$ instead of $X_i$ to leave room for
another abstract value $X_i^d$ later in the algorithm.

@@

Step~\ref{step:ascending} corresponds to the application of
path-focusing \cite{Monniaux_Gonnord_SAS11} to
the elements of $A$ until $A$ is empty. 
When $A$ becomes empty, it means the
invariant computation of the subgraph is finished. As proposed by
\citet{DBLP:conf/sas/GopanR07}, we can do some narrowing iterations.
These narrowing iterations allow to recover precision lost by
widening, \emph{before} computing and taking into account new feasible paths.
Thus, our technique combines both the advantages of \emph{Guided Static
Analysis} and \emph{Path-focusing}.

@@

For computing an inductive invariant over a subgraph, we use the
Path-focusing algorithm from \citet{Monniaux_Gonnord_SAS11} with special
treatment for self loops (line~\ref{alg=pf} in algorithm~\ref{algo:combined}).

@@

The difference with~\citet{Monniaux_Gonnord_SAS11} is that we do not
work on the entire transition graph but on a subset of it. Therefore we
conjoin the formula $f(p_i)$ with the actual set of working paths,
noted $P$, expressed as a Boolean formula, where the Boolean variables are the
\emph{reachability predicates} of the control points. We can easily construct
this formula from the binary decision diagram using dynamic programming, and
avoiding an exponentially sized formula. In other words, we force the SMT solver
to give us a path included in $P$.

@@

Guided static analysis from \citet{DBLP:conf/sas/GopanR07} yields, at line 5,
$\lstinline|x_old| \in (-\infty,+\infty)$.
Path-focusing \citep{Monniaux_Gonnord_SAS11} performs slightly
better, and finds $\lstinline|x_old| \in (-\infty,10000]$.
Now, let us see how our technique performs on this example.

@@

\citet{DBLP:conf/pldi/GulwaniZ10} proposed a technique for computing disjunctive invariants, by
distinguishing all the paths inside a loop. In
this section, we propose to improve this technique by using SMT queries to find
interesting paths, the objective being to avoid an explicit exhaustive
enumeration of an exponential number of paths.

@@

\item At each loop iteration of the original algorithm \citep{DBLP:conf/pldi/GulwaniZ10}, an image by each path inside the loop is computed for each disjunct of the invariant candidate.
Yet, many of these images may be redundant: for instance, if our invariant candidate is $(0 \leq x \leq 10 \land 0 \leq y \leq 1000) \lor (x < -10 \land y < -10)$, then there is no point enumerating paths whose image is included in this invariant candidate.
In our approach, we compute such an image only if it makes the resulting abstract value grow.
\end{itemize}

@@

\begin{eqnarray*}
g(p_i) & = & \rho \wedge b_i^s \wedge 
\displaystyle\bigwedge_{j \in P_R \atop j \neq i} \neg b_j^s  \\
 & & \wedge 
\displaystyle\bigvee_{1 \leq k \leq m_i} (d_k \wedge X_{i,k} \wedge \bigwedge_{l \neq k}
\neg d_l) \\
 & & \wedge
\displaystyle\bigvee_{j \in Succ(i)} 
(b_j^d \wedge \bigwedge_{1 \leq k \leq m_i} (\neg X_{j,k}))
\end{eqnarray*}

@@

The \emph{Update} function can classically assign to $X_{i,\sigma_i(j,k)}$ the
value $X_{i,\sigma_i(j,k)} \widening (X_{i,\sigma_i(j,k)} \sqcup
\tau_{i,k}(X_{i,j}))$, or can integrate the special treatment for self loops
proposed by \citet{Monniaux_Gonnord_SAS11}, with widening/narrowing sequence or
acceleration.

@@

We experimented with a heuristic
of dynamic construction of the $\sigma_i$ functions, adapted from
\citet{DBLP:conf/pldi/GulwaniZ10}. 
%\DM{Cette heuristique, je croyais que c'était nous qui l'avions trouvée,
%et qu'elle n'était pas dans Gulwani?}
%\JH{Non. Gulwani fait la même chose en un peu plus complique (voir commentaire
%suivant)}
%\MM{Du coup, la contribution~\ref{contr:disjunctive} devrait être
%  affaiblie un peu je suppose. Gulwani fait aussi du SMT-solving ?}
%\DM{Non, Gulwani énumère tous les chemins un par un à tous les coups.}
%\MM{OK, du coup j'ai ajouté une phrase à la fin de la section pour
%  rappeler notre contrib}
For each control point $p_i \in P_R$, we start with one single disjunct
($m_i = 1$) and define $\delta_i = 1$.
$M$ denotes an upper bound on the number of disjuncts per control point.

@@

%\MM{On peut peut-être citer le nom de l'outil (on garde PAGAI ?) ici.
%  L'intérêt d'avoir un nom, c'est que les gens peuvent en parler plus
%  facilement, et c'est plus facile de retrouver le papier origine d'un
%  outil quand il cite son nom}
%\DM{Dans la version finale, oui, mais pas dans une soumission anonymisée... suffit de Googlifier pagai sur imag.fr}
%
Our tool operates over LLVM bitcode \citep{LLVM_langref,Lattner:2004:LCF:977395.977673}, which is a target for several compilers, most notably Clang (supporting C and C++) and llvm-gcc (supporting C, C++, Fortran and Ada).
Abstract domains are provided by the APRON library \citep{DBLP:conf/cav/JeannetM09}, and include convex polyhedra (from the builtin Polka ``PK'' library), octagons, intervals, and linear congruences.
For SMT-solving, our analyzer uses Yices
\cite{DBLP:conf/cav/DutertreM06} or Microsoft Z3\cite{DBLP:conf/tacas/MouraB08}.

@@

\subsection{Analysis algorithm}
\label{sec:analysis-algorithm}
For each program, we distinguish a set $P_R = P_W$ of suitable widening points by a simple algorithm: for each procedure, compute the strongly connected components of its control-flow graph using Tarjan's algorithm; the targets of the back-edges of the depth-first search are added to~$P_R$. Note that the resulting set is not necessarily minimal, but is sufficient to disconnect all cycles --- more sophisticated techniques are discussed in e.g. \citet{BourdonclePhd}.

LLVM bitcode is in \emph{static single assignment} (SSA) form: a given scalar variable is given a value at a single syntactic point in the program. In concrete terms, an assignment \lstinline|x=2*x+1;| gets translated into a definition $x_2 = 2x_1+1$, with distinct variables $x_1$ and $x_2$ corresponding to the same original variable \lstinline|x| at different points in the program.
LLVM makes it easy to follow definition-use and use-definition chains: for a given variable (say, $x_2$) one can immediately obtain its definition (say, $2x_1+1$).
One may see conversion to SSA form as a static precomputation of some of the symbolic propagations proposed by \citet{DBLP:conf/vmcai/Mine06} to enhance the precision of analyses.

@@

This motivates a key implement decision of our tool: only those variables $v_1,\dots,v_n$ that are not defined by arithmetic operations are retained as coordinates in the abstract domain (e.g. as dimensions in polyhedra). A basic block of code therefore amounts to a \emph{parallel assignment} operation
$(v_1,\dots,v_n) \allowbreak\mapsto\allowbreak
(f_1(v_1,\dots,v_n), \allowbreak, \dots, \allowbreak
 f_n(v_1,\dots,v_n))$;
such operations are directly supported by APRON. This has three benefits:
(i) it limits the number of dimensions in the abstract values, since polyhedra libraries typically perform worse with higher dimensions;
(ii) the abstract operation for a single path in path-focusing methods also is a (large) parallel assignment;
(iii) as suggested by \citet{DBLP:conf/vmcai/Mine06}, this approach is more precise than running abstract operations for each program line separately:
for instance, for \lstinline|y=x; z=x-y;| with precondition $x \in [0,1]$, a line-by-line interval analysis obtains $y \in [0,1]$ and $z \in [-1,1]$ while our ``en bloc'' analysis symbolically simplifies $z = x - x = 0$ and thus $z \in [0,0]$.

@@

Our tool currently assumes that integer variables are unbounded mathematical integers ($\ZZ$) and floating-point variables are real (or rational) numbers. Techniques for sound analysis of bounded integers, including with wraparound, and of floating-point operations have been developed in e.g. the Astr\'ee system \citep{ASTREE_ESOP05,ASTREE_PLDI03}, but porting these techniques to our iteration schemes using SMT-solving requires supplemental work.

Our implementation of path-focusing currently does not use true acceleration techniques, as proposed by \citet{Monniaux_Gonnord_SAS11}. Instead, it simply runs widening and narrowing iterations on a single path.
The analysis is currently only forward, even though nothing in what we describe in this article is specific to forward analysis.

@@

\begin{figure}[h]
  \begin{center}
    \input{./techniques.tex}
  \end{center} 
  \vspace{-20pt}
  \caption{Comparison of the abstract values obtained by classical abstract
  interpretation (S), \emph{Guided Static
  Analysis} (G), \emph{Path-focused} technique (PF), our combined technique
  (G+PF), and its version with disjunctive invariants (DIS).
  The $\subsetneq$ bars gives the percentage of invariants stronger (more precise; smaller with respect to inclusion) with the left-side technique,
$\supsetneq$ the percentage of invariants stronger with the right-side technique,
and ``uncomparable'' gives the percentage of invariants that are uncomparable, i.e
neither greater nor smaller;
the code points where both invariants are equal make up the remaining percentage. See Tab.~\ref{tab:techniques} for details.}
  \label{fig:techniques}
\end {figure}

@@

Let us briefly comment the results given in more details in Table~\ref{tab:techniques} and Figure~\ref{fig:techniques}.
\emph{Guided Static Analysis} from \citet{DBLP:conf/sas/GopanR07} improves the
result of the classical Abstract Interpretation in $2.72\%$ of the control points
in $P_R$.
\emph{Path-focusing} from \citet{Monniaux_Gonnord_SAS11} gives
statistically better results, and finds better invariants than \emph{Guided
Static Analysis} in $9.89\%$ of the cases. However, it also loses precision in an
important number ($2.68\%$) of control points.
Finally, our combined technique gives the most promising results, since it is
statistically more precise than the other techniques.

@@

In cases where the formula is expressed in too rich a logic for the SMT-solver to deal with, a number of workarounds are possible, including:
(i) \emph{Linearization}, as per \citet{DBLP:conf/vmcai/Mine06}, which overapproximates nonlinear semantics by linear semantics.
(ii) Replacing the results of nonlinear operations by ``unknown''.
Neither is currently implemented in our tool.

@@

Finally, we evaluated the benefits of the improved version of the widening
operator for convex polyhedra from \citet{BagnaraHRZ05SCP}, compared to the
classical widening. We found that the improved version from
\citet{BagnaraHRZ05SCP} yields more precise invariants for $6.58\%$ of the
control points in $P_R$.

@@

% Cas a probleme:
% Initialisation (x,y) à {(1,0), (0,1)}
% puis boucle: si y = 0 alors x++
% fait un polyèdre qui finit par perdre y <= 1 et x >= 0

%\appendix
%\section{Appendix Title}

%This is the text of the appendix, if you need one.

%\acks

% Acknowledgments, if needed.

% We recommend abbrvnat bibliography style.

\phantomsection\addcontentsline{toc}{section}{References} 
\bibliographystyle{dmabbrvnat}
\bibliography{implicitization}
\end{document}

% Pour Emacs:
% Local Variables:
% reftex-cite-format: natbib
% End:

>>>>

\documentclass{llncs}

@@

\usepackage{amsmath}
\usepackage{mathcomp,amsfonts,amssymb,mathabx}
\usepackage[mathscr]{euscript}
\usepackage{algorithmic,algorithm}
%\usepackage{graphicx}
%\usepackage{color}

@@

\newcommand{\mytitle}{Succinct Representations for Abstract Interpretation}
\usepackage[pdftitle={\mytitle},pdfauthor={Julien Henry, David Monniaux, Matthieu Moy}]{hyperref}

@@

\graphicspath{{figures/}{gnuplot/}}

@@

\title{\mytitle}
\subtitle{Combined analysis algorithms and experimental evaluation}
\author{Julien Henry\inst{1}\inst{2} \and David Monniaux\inst{1}\inst{3} \and Matthieu Moy\inst{1}\inst{4}}

\institute{VERIMAG laboratory, Grenoble, France
\and Université Joseph Fourier
\and CNRS
\and Grenoble-INP}

\begin{document}
\maketitle

@@

@@

This article makes the following contributions:
\begin{enumerate}
\item We recast the guided static analysis technique from \cite{DBLP:conf/sas/GopanR07} on the expanded multigraph from \cite{Monniaux_Gonnord_SAS11}, considering entire paths instead of individual transitions, using SMT queries and binary decision diagrams.\label{contr:guided_multigraph} (See \S\ref{sec:guided_multigraph})
\item We improve the technique for obtaining disjunctive invariants from \cite{DBLP:conf/pldi/GulwaniZ10} by replacing the explicit exhaustive enumeration of paths by a sequence of SMT queries.\label{contr:disjunctive} (See \S\ref{sec:disjunctive})
\item We implemented these techniques, in addition to ``classical'' iterations and the original guided static analysis, inside a prototype static analyzer.
This tool uses the LLVM bitcode format \cite{Lattner:2004:LCF:977395.977673,LLVM_langref} as input, which can be produced by compilation from C, C++ and Fortran, enabling it to be run on many real-life programs.
With respect to abstract domains on numerical variables, it uses the APRON library \cite{DBLP:conf/cav/JeannetM09}, which supports a variety of abstract domains from which we can choose with minimal changes to our analyzer.

@@

The expected inductive invariant is $\lstinline|x_old| \in [-100000,100000]$, but classical abstract interpretation using intervals (or octagons or polyhedra) finds $\lstinline|x_old| \in (-\infty,+\infty)$~\cite{ASTREE_ESOP05}.
Let us briefly see why.

@@

We could use ``widening up to'' or ``widening with thresholds'', propagating the ``magic values'' $\pm 100000$ associated to \lstinline|x| into \mbox{\lstinline|x_old|,} but these syntactic approaches cannot directly cope with programs for which $\lstinline|x|  \in [-100000,+100000]$ is itself obtained by analysis.
The guided static analysis of \cite{DBLP:conf/sas/GopanR07} does not perform
better, and also obtains $\lstinline|x_old| \in (-\infty,+\infty)$.

@@

Following recent suggestions \cite{Gawlitza_Monniaux_ESOP11,Monniaux_Gonnord_SAS11}, we avoid this space explosion by keeping the second program implicit while simulating its analysis. This means we work on an implicitly represented transition multigraph (Fig.~\ref{fig:multigraph}); it is succinctly representated by the transition graph of the first program.
Our first contribution (\S\ref{sec:guided_multigraph}) is to recast the ``guided analysis'' from \cite{DBLP:conf/sas/GopanR07} on such a succinct representation of the paths in lieu of the individual transitions.
A similar explosion occurs in disjunctive invariant generation, following \cite{DBLP:conf/pldi/GulwaniZ10}; our second contribution (\S\ref{sec:disjunctive}) applies our implicit representation to their method.

\subsection{Guided static analysis}
\label{sec:guided}
\emph{Guided static analysis} was proposed by \cite{DBLP:conf/sas/GopanR07} as an improvement over classical upward Kleene iterations with widening.
Consider the following program, taken from \cite{DBLP:conf/sas/GopanR07}:
\lstinputlisting[label=lst:gopan_reps]{gopan_reps.c}

\begin{figure}
\begin{center}
\includegraphics[scale=0.75]{gopan_reps}
\end{center}
\caption{The invariant for Prog.~\ref{lst:gopan_reps}: the piecewise linear, solid line is the strongest invariant, the grayed polyhedron is its convex hull.}
\label{fig:gopan_reps_invariant} 
\end{figure}

@@

Gopan and Reps' idea is to analyze the first phase of the loop with a widening and narrowing sequence, and thus obtain $0 \leq x \leq 50 \land y = x$, and then analyze the second phase, finally obtaining invariant~\ref{eqn:triangle}; each phase is identified by the tests taken or not taken.

@@

\cite{Monniaux_Gonnord_SAS11}'s \emph{path-focusing} technique
distinguishes the different paths in the program in order to avoid loss of
precision due to merge operations. Since the number of paths may be exponential,
the technique keeps them implicit and computes them when needed using
SMT-solving.
The (accelerated) Kleene iterations (\S\ref{sec:static_analysis}) are computed over a reduced multigraph instead of the classical transition graph.

@@

Guided static analysis, as proposed by \cite{DBLP:conf/sas/GopanR07}, applies to the transition
graph of the program. We now present a new technique applying this analysis on the implicit
multigraph from \cite{Monniaux_Gonnord_SAS11}, thus avoiding control flow merges with
unfeasible paths.
In this section, we use the same notations as \ref{sec:path_focusing}, except
that we call the abstract values $X_i^s$ instead of $X_i$ to leave room for
another abstract value $X_i^d$ later in the algorithm.

@@

Step~\ref{step:ascending} corresponds to the application of
path-focusing \cite{Monniaux_Gonnord_SAS11} to
the elements of $A$ until $A$ is empty. 
When $A$ becomes empty, it means the
invariant computation of the subgraph is finished. As proposed by
\cite{DBLP:conf/sas/GopanR07}, we can do some narrowing iterations.
These narrowing iterations allow to recover precision lost by
widening, \emph{before} computing and taking into account new feasible paths.
Thus, our technique combines both the advantages of \emph{Guided Static
Analysis} and \emph{Path-focusing}.

@@

For computing an inductive invariant over a subgraph, we use the
Path-focusing algorithm from \cite{Monniaux_Gonnord_SAS11} with special
treatment for self loops (line~\ref{alg=pf} in algorithm~\ref{algo:combined}).

@@

The difference with~\cite{Monniaux_Gonnord_SAS11} is that we do not
work on the entire transition graph but on a subset of it. Therefore we
conjoin the formula $f(p_i)$ with the actual set of working paths,
noted $P$, expressed as a Boolean formula, where the Boolean variables are the
\emph{reachability predicates} of the control points. We can easily construct
this formula from the binary decision diagram using dynamic programming, and
avoiding an exponentially sized formula. In other words, we force the SMT solver
to give us a path included in $P$.

@@

Guided static analysis from \cite{DBLP:conf/sas/GopanR07} yields, at line 5,
$\lstinline|x_old| \in (-\infty,+\infty)$.
Path-focusing \cite{Monniaux_Gonnord_SAS11} performs slightly
better, and finds $\lstinline|x_old| \in (-\infty,10000]$.
Now, let us see how our technique performs on this example.

@@

\cite{DBLP:conf/pldi/GulwaniZ10} proposed a technique for computing disjunctive invariants, by
distinguishing all the paths inside a loop. In
this section, we propose to improve this technique by using SMT queries to find
interesting paths, the objective being to avoid an explicit exhaustive
enumeration of an exponential number of paths.

@@

\item At each loop iteration of the original algorithm \cite{DBLP:conf/pldi/GulwaniZ10}, an image by each path inside the loop is computed for each disjunct of the invariant candidate.
Yet, many of these images may be redundant: for instance, if our invariant candidate is $(0 \leq x \leq 10 \land 0 \leq y \leq 1000) \lor (x < -10 \land y < -10)$, then there is no point enumerating paths whose image is included in this invariant candidate.
In our approach, we compute such an image only if it makes the resulting abstract value grow.
\end{itemize}

@@

\begin{equation*}
g(p_i) = \rho \wedge b_i^s \wedge 
\displaystyle\bigwedge_{j \in P_R \atop j \neq i} \neg b_j^s 
 \wedge 
\displaystyle\bigvee_{1 \leq k \leq m_i} (d_k \wedge X_{i,k} \wedge \bigwedge_{l \neq k}
\neg d_l)
\wedge
\displaystyle\bigvee_{j \in Succ(i)} 
(b_j^d \wedge \bigwedge_{1 \leq k \leq m_i} (\neg X_{j,k}))
\end{equation*}

@@

The \emph{Update} function can classically assign to $X_{i,\sigma_i(j,k)}$ the
value $X_{i,\sigma_i(j,k)} \widening (X_{i,\sigma_i(j,k)} \sqcup
\tau_{i,k}(X_{i,j}))$, or can integrate the special treatment for self loops
proposed by \cite{Monniaux_Gonnord_SAS11}, with widening/narrowing sequence or
acceleration.

@@

We experimented with a heuristic
of dynamic construction of the $\sigma_i$ functions, adapted from
\cite{DBLP:conf/pldi/GulwaniZ10}. 
%\DM{Cette heuristique, je croyais que c'était nous qui l'avions trouvée,
%et qu'elle n'était pas dans Gulwani?}
%\JH{Non. Gulwani fait la même chose en un peu plus complique (voir commentaire
%suivant)}
%\MM{Du coup, la contribution~\ref{contr:disjunctive} devrait être
%  affaiblie un peu je suppose. Gulwani fait aussi du SMT-solving ?}
%\DM{Non, Gulwani énumère tous les chemins un par un à tous les coups.}
%\MM{OK, du coup j'ai ajouté une phrase à la fin de la section pour
%  rappeler notre contrib}
For each control point $p_i \in P_R$, we start with one single disjunct
($m_i = 1$) and define $\delta_i = 1$.
$M$ denotes an upper bound on the number of disjuncts per control point.

@@

%\MM{On peut peut-être citer le nom de l'outil (on garde PAGAI ?) ici.
%  L'intérêt d'avoir un nom, c'est que les gens peuvent en parler plus
%  facilement, et c'est plus facile de retrouver le papier origine d'un
%  outil quand il cite son nom}
%\DM{Dans la version finale, oui, mais pas dans une soumission anonymisée... suffit de Googlifier pagai sur imag.fr}
%
Our tool operates over LLVM bitcode \cite{LLVM_langref,Lattner:2004:LCF:977395.977673}, which is a target for several compilers, most notably Clang (supporting C and C++) and llvm-gcc (supporting C, C++, Fortran and Ada).
Abstract domains are provided by the APRON library \cite{DBLP:conf/cav/JeannetM09}, and include convex polyhedra (from the builtin Polka ``PK'' library), octagons, intervals, and linear congruences.
For SMT-solving, our analyzer uses Yices
\cite{DBLP:conf/cav/DutertreM06} or Microsoft Z3\cite{DBLP:conf/tacas/MouraB08}.

@@

\subsection{Analysis algorithm}
\label{sec:analysis-algorithm}
For each program, we distinguish a set $P_R = P_W$ of suitable widening points by a simple algorithm: for each procedure, compute the strongly connected components of its control-flow graph using Tarjan's algorithm; the targets of the back-edges of the depth-first search are added to~$P_R$. Note that the resulting set is not necessarily minimal, but is sufficient to disconnect all cycles --- more sophisticated techniques are discussed in e.g. \cite{BourdonclePhd}.

LLVM bitcode is in \emph{static single assignment} (SSA) form: a given scalar variable is given a value at a single syntactic point in the program. In concrete terms, an assignment \lstinline|x=2*x+1;| gets translated into a definition $x_2 = 2x_1+1$, with distinct variables $x_1$ and $x_2$ corresponding to the same original variable \lstinline|x| at different points in the program.
LLVM makes it easy to follow definition-use and use-definition chains: for a given variable (say, $x_2$) one can immediately obtain its definition (say, $2x_1+1$).
One may see conversion to SSA form as a static precomputation of some of the symbolic propagations proposed by \cite{DBLP:conf/vmcai/Mine06} to enhance the precision of analyses.

@@

This motivates a key implement decision of our tool: only those variables $v_1,\dots,v_n$ that are not defined by arithmetic operations are retained as coordinates in the abstract domain (e.g. as dimensions in polyhedra). A basic block of code therefore amounts to a \emph{parallel assignment} operation
$(v_1,\dots,v_n) \allowbreak\mapsto\allowbreak
(f_1(v_1,\dots,v_n), \allowbreak, \dots, \allowbreak
 f_n(v_1,\dots,v_n))$;
such operations are directly supported by APRON. This has three benefits:
(i) it limits the number of dimensions in the abstract values, since polyhedra libraries typically perform worse with higher dimensions;
(ii) the abstract operation for a single path in path-focusing methods also is a (large) parallel assignment;
(iii) as suggested by \cite{DBLP:conf/vmcai/Mine06}, this approach is more precise than running abstract operations for each program line separately:
for instance, for \lstinline|y=x; z=x-y;| with precondition $x \in [0,1]$, a line-by-line interval analysis obtains $y \in [0,1]$ and $z \in [-1,1]$ while our ``en bloc'' analysis symbolically simplifies $z = x - x = 0$ and thus $z \in [0,0]$.

@@

Our tool currently assumes that integer variables are unbounded mathematical integers ($\ZZ$) and floating-point variables are real (or rational) numbers. Techniques for sound analysis of bounded integers, including with wraparound, and of floating-point operations have been developed in e.g. the Astr\'ee system \cite{ASTREE_ESOP05,ASTREE_PLDI03}, but porting these techniques to our iteration schemes using SMT-solving requires supplemental work.

Our implementation of path-focusing currently does not use true acceleration techniques, as proposed by \cite{Monniaux_Gonnord_SAS11}. Instead, it simply runs widening and narrowing iterations on a single path.
The analysis is currently only forward, even though nothing in what we describe in this article is specific to forward analysis.

@@

\begin{figure}[h]
  \begin{center}
    \input{gnuplot/techniques.tex}
  \end{center} 
  \vspace{-20pt}
  \caption{Comparison of the abstract values obtained by classical abstract
  interpretation (S), \emph{Guided Static
  Analysis} (G), \emph{Path-focused} technique (PF), our combined technique
  (G+PF), and its version with disjunctive invariants (DIS).
  The $\subsetneq$ bars gives the percentage of invariants stronger (more precise; smaller with respect to inclusion) with the left-side technique,
$\supsetneq$ the percentage of invariants stronger with the right-side technique,
and ``uncomparable'' gives the percentage of invariants that are uncomparable, i.e
neither greater nor smaller;
the code points where both invariants are equal make up the remaining percentage. See Tab.~\ref{tab:techniques} for details.}
  \label{fig:techniques}
\end {figure}

@@

Let us briefly comment the results given in more details in Table~\ref{tab:techniques} and Figure~\ref{fig:techniques}.
\emph{Guided Static Analysis} from \cite{DBLP:conf/sas/GopanR07} improves the
result of the classical Abstract Interpretation in $2.72\%$ of the control points
in $P_R$.
\emph{Path-focusing} from \cite{Monniaux_Gonnord_SAS11} gives
statistically better results, and finds better invariants than \emph{Guided
Static Analysis} in $9.89\%$ of the cases. However, it also loses precision in an
important number ($2.68\%$) of control points.
Finally, our combined technique gives the most promising results, since it is
statistically more precise than the other techniques.

@@

In cases where the formula is expressed in too rich a logic for the SMT-solver to deal with, a number of workarounds are possible, including:
(i) \emph{Linearization}, as per \cite{DBLP:conf/vmcai/Mine06}, which overapproximates nonlinear semantics by linear semantics.
(ii) Replacing the results of nonlinear operations by ``unknown''.
Neither is currently implemented in our tool.

@@

Finally, we evaluated the benefits of the improved version of the widening
operator for convex polyhedra from \cite{BagnaraHRZ05SCP}, compared to the
classical widening. We found that the improved version from
\cite{BagnaraHRZ05SCP} yields more precise invariants for $6.58\%$ of the
control points in $P_R$.

@@

\phantomsection\addcontentsline{toc}{section}{References} 
\bibliographystyle{splncs03}
\bibliography{implicitization}
\end{document}
